---
title: "Survival analysis for the Stanford Heart Transplant Data"
author: "Oleg Kufirin"
date: "19/03/2021"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(comment = NA)
```



## Dataset

[Link to the interactive Shiny app](https://oleg-kufirin.shinyapps.io/HTx_Survival_Analysis/)

<div style="text-align: justify">
Time-dependent variables can be used to model the effects of subjects 
transferring from one treatment group to another. One example of the need for 
such strategies is the Stanford heart transplant program. Patients are accepted 
if physicians judge them suitable for heart transplant. Then, when a donor 
becomes available, physicians choose transplant recipients according to various 
medical criteria. A patient's status can be changed during the study from 
waiting for a transplant to transplant recipient. 

Crowley and Hu (J. Amer. Statist. Assoc., 1977) describe a program at Stanford 
University where 103 patients were registered to recieve a heart transplant and 
were followed until death or censorship. Of these, 65 received transplants 
during follow-up, whereas 38 did not.
</div>

Here is the head of the dataset imported from *p3state.msm* package:

```{r warning=FALSE, message=FALSE}
library(survival)
library(survminer)
library(p3state.msm)
library(Hmisc)
data(heart2)
head(heart2)
```

[Reference to the dataset description](https://rdrr.io/cran/p3state.msm/man/heart2.html)



## Study question

<div style="text-align: justify">
In this project we would like to assess the impact of the following
characteristics of patients from the heart transplant waiting list on their 
survival. 

- Transplant event. Do patients with transplanted hearts live longer?
- Age group. Do certain age groups have better or worse prognosis?
- Year group. Is there any difference in survival of patients from different
groups of time of acceptance into the program after 1 Nov 1967?
- Surgery. Does having a prior bypass surgery lead to better or worse prognosis?

To do this we will apply Kaplan-Meier estimate and Cox Proportional Hazard Model.
The app allows manipulation of model parameters and visualises survival curves.

#### Data preparation
Variables *times1* and *times2* were ignored for the analysis. Variable
*delta* was renamed to *Transplant*. Variable *age* was brought to the original
scale and categorised to *Age Group*. Variable *year* was also categorised 
to *Year Group*.
</div>



## Kaplan-Meier Survival Estimate

<div style="text-align: justify">
On the first tab of the app a user can see the Kaplan-Meier Survival Curve and 
the Cumulative Hazard Curve for chosen variable of interest. All variables are
categorical and described in the data dictionary at the bottom of the left panel. 
A user can choose factor of interest from Stratification Factor drop-down list.

Below is an example of a KM curve for Age Group variable. As it can be seen
younger patients below 40 have better survival outcome given they are 
accepted in the waiting list. This is not a surprising result but it shows
how other variables can be explored in a similar fashion.
</div>

```{r warning=FALSE, message=FALSE, out.width="45%", fig.align='center'}
# -------------------------------------------------------------
# times1    time of transplant/censoring time
# delta     transplant indicator
# times2    time to death since the transplant/censoring time
# time      times1 + times2
# status    censoring indicator: dead=1, alive=0
# age       age-48 years
# year      year of acceptance; in years after 1 Nov 1967
# surgery   prior bypass surgery; 1=yes
# -------------------------------------------------------------

# original age column was centred at 48, go back to the original scale
heart2$age <- heart2$age + 48
# add new column for age groups
heart2$age.group <- cut2(heart2$age, c(40,50))
# add new column for age year of inclusion into WL groups
heart2$year.group <- cut2(heart2$year, c(2,5))

colnames(heart2)[2] <- "transplant"
heart2$transplant <- as.factor(heart2$transplant)
heart2$surgery <- as.factor(heart2$surgery)


KaplanMeier <- function(parameter) {
  expr = paste("survfit(Surv(time, status) ~ ", parameter, ", data = heart2)")
  fit <- eval(parse(text = expr))
  # Survival curve
  surv_curve <- ggsurvplot(
                            fit,
                            title = "Kaplan-Meier survival curve",
                            pval = TRUE,
                            conf.int = TRUE,
                            # conf.int.style = "step",
                            xlab = "Time in days",
                            break.time.by = 200,
                            ggtheme = theme_light(),
                            risk.table = "abs_pct",
                            risk.table.y.text.col = T,
                            risk.table.y.text = FALSE,
                            ncensor.plot = TRUE,
                            surv.median.line = "hv",
                            xlim = c(0, 1800)
                          )

  # Cumulative hazard
  cumul_haz <- ggsurvplot(
                            fit,
                            title = "Cumulative hazard curve",
                            conf.int = TRUE,
                            risk.table.col = "strata",
                            ggtheme = theme_bw(),
                            fun = "cumhaz"
                        )
  return(list(survival = surv_curve, hazard = cumul_haz))
}

graphs <- KaplanMeier("age.group")
graphs$survival
```

[Reference about Kaplan-Meier Survival Estimate](http://www.sthda.com/english/wiki/survival-analysis-basics)



## Cox Proportional-Hazards Model

<div style="text-align: justify">
On the second tab a user can see a survival curve built with 
Cox Proportional-Hazards Model, model summary and the predicted curve based on 
the user's input. Regressors that a user wants to include in the model can be 
selected with the tick-boxes on the left panel. Model summary is given below 
the curve and contains useful information such as coefficient values and results 
of statistical tests. Survival curve for a new patient will be drawn at the 
bottom where input values are specified on the left panel under 'Prediction' 
section. Only regressors included in the model will have an effect on prediction 
despite other existing input values.

Below is an example of a Cox PH curve of a univariate model for Transplant
variable.
</div>

```{r warning=FALSE, message=FALSE, out.width="30%", fig.align='center'}
res.cox = coxph(Surv(time, status) ~ transplant, data = heart2)
df <- with(heart2, data.frame(transplant = c('0', '1')))

ggsurvplot(survfit(res.cox, newdata = df), data = heart2,
           ggtheme = theme_minimal(), legend.labs = c("Transplant=No",
                                                      "Transplant=Yes"))
```

[Reference about Cox Proportional-Hazards Model](http://www.sthda.com/english/wiki/cox-proportional-hazards-model)



